---
title: "Final Project"
author: "Muhammed Abubaker Syed, Vy Nguyen, Lanlie Nguyen, Spencer Wolf"
chunk_output_type: inline
output: html_document
---

```{r}
library(ggplot2)
library(tidyverse)
library(lubridate)
library(dplyr)
library(tibble)
```

## Importing the Indian Premiere League dataset. The dataset contains all the matches ever played in the game starting 2008. 

```{r}
data <- read.csv(url("https://raw.githubusercontent.com/SYEDM-USA/MATH-3430/main/IPL%20Matches%202008-2020.csv"))
data
```

## Untidy -ing the tidy dataset

```{r}
mangled_data = gather(data, key=Field, value=FieldValue, city, venue, neutral_venue, date) %>%
  gather(key=TeamNumber, value=Team, team1, team2) %>%
  unite(umpire_team, umpire1, umpire2, sep="/") %>%
  rename(t=toss_decision)
mangled_data
```

## Tidying the dataset

```{r}
Data = spread(mangled_data, key=Field, value=FieldValue) %>%
  spread(key=TeamNumber, value=Team) %>%
  separate(umpire_team, c("umpire1", "umpire2"), sep="/") %>%
  rename(toss_decision = t)
head(Data,5)
ncol(Data)
```


## Analysis of the Dataset. 
Some questions which are always asked when dealing with sports data is performance of each team over the years. 
Drop columns
`id`, `city`,`neutral_venue`,`toss_decision`,`eliminator`,`method`
```{r}
Final<-select(Data,-c('id','city','neutral_venue','toss_decision','eliminator','method'))
head(Final,5)
ncol(Final)
Final
```

1. Show the EDA for the top 10 umpire combinations.
```{r}
Umpires<-select(Final,c('umpire1','umpire2'))%>%
  group_by(umpire1,umpire2)%>%
  summarise(Games_umpired = n())%>%
  arrange(desc(Games_umpired))%>%
  head(10)
Umpires
```
2. Who are the top 10 most popular umpires, show a bar plot of the no. of games they umpired
```{r}
umpire1_<- select(Final,c('umpire1'))%>%
  rename('Umpire'='umpire1')
umpire2_ = select(Final,c('umpire2'))%>%
  rename('Umpire'='umpire2')
Umpires_s <- bind_rows(umpire1_,umpire2_)%>%
  group_by(Umpire)%>%
  summarise(Games_Umpired = n())%>%
  arrange(desc(Games_Umpired))%>%
  head(10)
Umpires_s
```
3. Top 10 teams with the best win rate
```{r}
Win_team<- Final%>%
  rename('Team'='winner')%>%
  group_by(Team)%>%
  summarise(wins = n())%>%
  arrange(desc(wins))%>%
  drop_na()
team1_<- select(Final,c('team1'))%>%
  rename('Team'='team1')
team2_ = select(Final,c('team2'))%>%
  rename('Team'='team2')
Team_s <- bind_rows(team1_,team2_)%>%
  group_by(Team)%>%
  summarise(Games_played = n())
merge(Team_s,Win_team)%>%
  mutate(Win_rate = Games_played/wins)%>%
  arrange(desc(Win_rate))
```
4. Which stadiums are used most?
```{r}
Final%>%
  group_by(venue)%>%
  summarise(no_of_games_played = n())%>%
  arrange(desc(no_of_games_played))
```

5. Which players have won the most man of the match awards
```{r}
Final%>%
  group_by(player_of_match)%>%
  summarise(Games_as_Man_of_Match = n())%>%
  arrange(desc(Games_as_Man_of_Match))
```
6. Which team is the gambler?
```{r}
Final%>%
  group_by(toss_winner)%>%
  summarise(Toss_wins = n())%>%
  arrange(desc(Toss_wins))%>%
  rename('Teams'=toss_winner)
```

7. What game has the highest score difference between teams?
```{r}
Final%>%
  filter(result=="runs")%>%
  arrange(desc(result_margin))
```

8. Which teams choose the right option after toss to win the game most number of times?
```{r}
Final%>%
  filter(toss_winner==winner)%>%
  group_by(winner)%>%
  summarise(No_of_games_choose_the_right_option=n())%>%
  arrange(desc(No_of_games_choose_the_right_option))
```

9. Scatter plor, showing the average of result wicket or runs for each of the 15 teams
```{r}
Final%>%
  select(c('date','winner','result','result_margin'))%>%
  spread(key= 'result', value = 'result_margin')%>%
  select(-c('<NA>',tie,'date'))%>%
  group_by(winner)%>%
  summarise(avg_runs = mean(runs,na.rm = T),
            avg_wickets = mean(wickets,na.rm = T))%>%
  gather(c(avg_runs,avg_wickets),key = type, value=avg)%>%
  drop_na()%>%
  ggplot(aes(winner,avg,fill=type))+
  geom_bar(stat="identity",position = "dodge")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

10. Champions by year
```{r}
helper<- Final%>%mutate(date = dmy(date))
Champions<-Final%>%
  mutate(Season = year(dmy(date)))%>%
  group_by(Season)%>%
  summarise(date = max(dmy(date)))%>%
  left_join(helper)%>%
  select(c(Season,winner, player_of_match, venue))
Champions
```

Team with most Final wins
```{r}
Champions%>%
  group_by(winner)%>%
  summarise(No_of_wins = n())%>%
  arrange(desc(No_of_wins))
```

#if x team bats secoindand wins, then result is in wickets 
#if x team bats first  and wins , the rresult is in runs 
